---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1.** 26 departments have offered courses with "Climate" in the title and at least one seat. Atmospheric and Oceanic Science has offered the most, with 68 classes.

```{r}
# copy data set
climate_courses <- umd_courses |> 
  filter(
# filter data to detect "climate" in course title
    str_detect(title, "Climate")
# AND filter data for courses that have at least one seat
    & seats >= 1
  ) |> 
# group data by department
  group_by(department) |> 
# count number of filtered classes per department
  summarize(
      count_id = n()
  ) |> 
# sort data by counted filtered classes per department, in descending order
arrange(desc(count_id))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? **A2.** Germanic Studies is the most interesting to me, as I'm assuming this is a historical/linguistic-based discipline, and I'm wondering what they'd be studying climate for.

**Q3** Which departments have the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? How are the top five departments different in each result?

**A3.**

Most courses in Fall 2023

1.  Business and Management

2.  Music

3.  English

4.  Teaching and Learning, Policy and Leadership

5.  Information Studies

Most syllabi posted

1.  Business and Management

2.  Economics

3.  Music

4.  Biological Sciences Program

5.  Information Studies

```{r}
# copy data set
fall2023_courses <- umd_courses |> 
# filter data to fall 2023
  filter(
    term == 202308
  ) |> 
# group data by department
  group_by(department) |> 
# count number of classes per department
  summarize(
    count_id = n()
  ) |> 
# arrange by counted classes
arrange(desc(count_id))
# show top
head(fall2023_courses)

# to answer second question
fall2023_syllabi <- umd_courses |> 
# filter data to fall 2023 AND at least one syllabus counted
  filter(
    term == 202308
    & syllabus_count >= 1
  ) |> 
# group data by department
  group_by(department) |> 
# count syllabus_count per department
  summarize(
    count_syllabus_count = n()
  ) |> 
# arrange by counted syllabus_count by department
arrange(desc(count_syllabus_count))
# show top
head(fall2023_syllabi)

```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as the same dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name). Write a sentence describing the pattern of spending over time. **A4.** There's waves of spending and no spending from 2021-2022, with the largest amounts being spent leading up to the 2022 election.

```{r}
# alter data type to date, not chr
maryland_expenses <- maryland_expenses |> 
  mutate(
    expenditure_date=ymd(expenditure_date)
  )

# answered the rest of the question on this copy
maryland_expenses_work <- maryland_expenses |> 
# create month column
  mutate(
    month = floor_date(expenditure_date, "month")
  ) |> 
# group data by month
  group_by(month) |> 
# count total spent per month
  summarize(
    total_amount = sum(amount)
  )

```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for the others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). **A5** According to the code, the campaign spent over $77 million in Maryland and nothing out of state, but that doesn't make sense to me. I'm expecting some expenditures coming from out of state, with most of the money still being concentrated in-state.

```{r}
# copy data set
maryland_expenses_location <- maryland_expenses |> 
# create column
  mutate(
    spending_location = case_when(
# if address is in MD, assign value in-state
      str_detect(address, " Maryland ") ~"In-state",
# if not, assign value "out of state"
      .default = "Out of state"
    )
  ) |> 
# group expenditures by location
  group_by(spending_location) |> 
# sum amounts by location
  summarize(
    total_spent = sum(amount)
  )
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.** Music, with 59 classes

```{r}
# question: which department has the most classes with less than or equal to 10 seats, in the  fall 2023 semester?

small_courses <- umd_courses |> 
# filter data to only fall 2023 semester, with classes with less than/equal to 10 seats
  filter(
    term == 202308
    & seats <= 10
  ) |> 
# group data by department
  group_by(department) |> 
# count filtered classes per department
  summarize(
    count_id = n()
  ) |> 
# arrange by counted classes
  arrange(desc(count_id))

# print top results
head(small_courses)
```
