---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Hello puppy!")
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/code/data_journalism_2023_spring")
```

### How to install and load the tidyverse

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

### How to read in a CSV file from the Internet

CSV files are spreadsheets.

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```

### How to set a variable

```{r}
variable_name <- value
variable_name <- command()
```

### How to count rows

You use the `n()` function to count rows, like so:

```{r}
dataframe %>% 
  group_by(some_column) %>% 
  summarize(count = n())
```

## Functions

### glimpse()

Lays out variables (columns of spreadsheet) and the data type of each variable.

Data types: - char = character, or a string of text - dbl = number - logical = boolean - date - datetime

### head()

Gives you the first 6 rows of data.

### group_by()

Groups data by **column heads.** Can group by multiple columns.

### summarize()

Holding function to compute math.

"how many?" = n() or counting

"how much?" = sum()

### arrange()

Outputs arrangement of data

### summary()

Provides length, data type and (if data type is numeric) quartiles. Good way to see if there's any interesting points in the data.

### colnames()

Prints out column names of dataset.

### filter()

Creates a subset of the data, by altering what **rows** you see. Try to do this first before other functions.

```{r}
filter(operator == criteria)
# for numeric filters, can use > or <
# use "" to distinguish character criteria. not needed for numbers
# use str_detect(column_head, "criteria") to filter using a part of a string
```

### select()

Creates a subset of the data, by altering what **columns** you see. Used less compared to filter(). You will have the same number of records when you are done, but fewer columns of data to work with.

### mutate()

Alters an existing column or creates a new one. mutate(column = change) Can populate things, create things or alter data types.

### rename()

Renames header rows.

### bind_rows()

Combines data sets IF the column structure is the same (order can be variable, but needs to be same syntax). Should be saved to a new data frame. Makes a data set longer (more rows.)

```{r}
combined <- originalset |> 
  bind_rows(list(dataframe1, dataframe2))

```

### left_join()

Combines data sets and displays every row even if there's no match. Best to do first.

```{r}
# data with more records should be first
# take first data frame, join with second data frame, using common column name as the key
combined <- dataframe1 |> 
  left_join(dataframe2, join_by("variable"))

```

### inner_join()

Combines data sets and displays only the overlap.

### operators

```{r}
# <- variable assignment
# = assignment
# == equality
# & boolean, and. both conditions need to be true for it to be included
# | boolean, or. any of the conditions can be true and it will be included
```

## Random notes

-   Basic formula of what to write:

```{r}
data |>
  group_by(COLUMN NAME) |>
  summarize(
    VARIABLE NAME = AGGREGATE FUNCTION(COLUMN NAME)
    )
```

-   Tables are sorted in ascending order. To change, use arrange(desc(variable))
-   When it comes to filters, OR is additive/inclusive; AND is restrictive/exclusive.
-   ! = boolean/exclusionary filter:
```{r}
# filter for payee names that are NOT n/a values
filter(!is.na(payee_name))
# another option is
drop_na(column_name)
```