---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Hello puppy!")
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/code/data_journalism_2023_spring")
```

### How to install and load the tidyverse

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

### How to read in a CSV file from the Internet

CSV files are spreadsheets.

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```

### How to set a variable

```{r}
variable_name <- value
variable_name <- command()
```

### How to count rows

You use the `n()` function to count rows, like so:

```{r}
dataframe %>% 
  group_by(some_column) %>% 
  summarize(count = n())
```

## Commands

### glimpse()

Lays out variables (columns of spreadsheet) and the data type of each variable.

Data types:
-   char = character, or a string of text
-   dbl = number
-   logical = boolean
-   date
-   datetime

### head()

Gives you the first 6 rows of data.

## Random notes
- Tables are sorted in ascending order. To change, use arrange(desc(variable))
