title: "Reverse Engineering Project"
author: "Student names here"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story, [L.A. is slammed with record costs for legal payouts](https://www.latimes.com/local/lanow/la-me-ln-city-payouts-20180627-story.html)

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}

library(tidyverse)
library(lubridate)
library(janitor)

```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# set working directory as reverse-engineering-la-times and loan required data
la_payouts <- read_csv("data/payouts/payouts.csv")
la_departments <- read_csv("data/departments.csv")
la_casetypes <- read_csv("data/casetypes.csv")

```

```{r}

# clean data

la_payouts_cleaned <- la_payouts
# create column names from first row
colnames(la_payouts_cleaned) <- la_payouts[1,]
# remove first row
la_payouts_cleaned <- la_payouts_cleaned[-1,]


#la_payouts_cleaned <- la_payouts_cleaned |> mutate(AMOUNT=as.numeric(AMOUNT))

la_payouts_cleaned

la_payouts_cleaned |> 
  clean_names() |> 
  mutate(
    department = str_to_upper(department),
    case_type = str_to_upper(case_type)
  )

```


```{r}
#la_payouts_cleaned <- la_payouts_cleaned |>
  #mutate(
   # state = case_when(
       # str_detect("AMOUNT", "$") ~ "",
       # str_detect("AMOUNT", ",") ~ "",
       # .default = NA
    #  )
 # )

#la_payouts_cleaned
```
```{r}
la_payouts_cleaned["AMOUNT"] <- lapply(la_payouts_cleaned["AMOUNT"], function(x) as.numeric(gsub("[,$]", "", x)))

la_payouts_cleaned
```
