---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
# already set API key in pre-lab
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2021, "acs5")
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 

**A1** From 2016-2021, the Maryland jurisdiction with the lowest median age was estimated to be Baltimore City, at 35.7 years.

```{r}
# create dataframe that median age per county in maryland, using ACS 2021 5-year estimate
md <- get_acs(geography = "county",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2021)

md |> 
  arrange(estimate)

```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 
**A2** Talbot and Worcester counties are the only Maryland counties that have a median age of at least 50.

```{r}
md |> 
  filter(estimate >= 50) 

```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1. Load the Montgomery County 911 overdose calls data.
2. Create a new dataframe with the total number of calls for each zip code.
3. Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).
4. Get from the ACS data from all Maryland zip codes using that variable.
5. Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
6. Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?

**A3** The zip code that has the highest rate of calls in Montgomery County is 20877, which encompasses part of Gaithersburg and has 1.7 calls per 1,000 people. The zip code that has the lowest rate of calls is 20707, which has 0.04 calls per 1,000 people. However, a majority of this zipcode encompasses land in Laurel, which is in Prince George's County, not Montgomery County.

I also want to note that one listed zip code in the overdose call data, 20889, did not have accompanying demographic data from the Census Bureau. Looking up the zip code online reveals that it encompasses a residential community, so I'm wondering why there is a gap in the data.

```{r}
# load moco overdose data
moco_911 <- read_csv("data/montgomery_2022_overdoses.csv") |> 
# make zip char
  mutate(
    zip = as.character(zip)
  )
# load age data by zip in md
md_age <- get_acs(geography = "zcta",
                  variables = c(adults_18_and_over = "B09021_001"),
                  state = "MD",
                  year = 2019)
  
# copy data frame
moco_911_zip <- moco_911 |> 
# group by zip
  group_by(zip) |> 
# count calls per zip
  summarize(
    count_calls = n()
  )

# create new data frame
moco_911_with_age <- moco_911 |> 
# join moco_911 with md_age
  left_join(md_age, by=c("zip" = "GEOID")) |> 
# clean up
  select(-NAME, -variable) |> 
  rename(adults_18_and_over = estimate)

moco_911_with_age_percapita <- moco_911_with_age |> 
# group by zip
  group_by(zip, adults_18_and_over, moe) |>
# count calls per zip
  summarize(
    count_calls = n()
  ) |> 
# add per capita column
  mutate(
    calls_per_capita = count_calls/adults_18_and_over * 1000
  ) |> 
arrange(desc(calls_per_capita))
moco_911_with_age_percapita
  
```

**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4** I do not think that a proper, equitable comparison could be made between the two zip codes (if the intention is to analyze solely Montgomery County's overdose calls), because they are located in different counties. Moreover, the zip codes are somewhat similar demographically. 

The Gaithersburg zip code is a majority minority community, that is 42% Hispanic. The median household income is estimated at \$77,264 and around 15% of people live below the poverty line. The median age is 37.6.

The Laurel zip code has similar numbers. It is a majority minority community that is 45% Black. The median household income is estimated at \$91,315 and around 8.4% of people live below the poverty line. The median age is 38.

**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5** Yes, taking into account the margins of error, the per capita rates still result in 20877 having the highest rate of calls and 20707 having the lowest rate of calls. This does give more confidence to my answers in Q3.

```{r}
# copy data frame
moco_911_with_age_percapita_moe <- moco_911_with_age_percapita |> 
# add columns
  mutate (
# calculate max pop
    adults_max = adults_18_and_over + moe,
# calculate calls per max capita
    calls_per_capita_max = count_calls / adults_max * 1000,
# calculate min pop 
    adults_min = adults_18_and_over - moe,
# calculate calls per min capita
    calls_per_capita_min = count_calls / adults_min * 1000
  ) |> 
# arrange(desc(calls_per_capita_max))
arrange(desc(calls_per_capita_min))

```

