---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)
```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
md_expenses <- read_csv("data/maryland_expenses.csv") |> 
# clean up column names
  clean_names() |> 
# fix date column
  mutate(
    expenditure_date = mdy(expenditure_date)
  )

```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

1.  Create a dataframe of Media expenses where the name of the payee is not NA.
2.  Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
3.  Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
4.  When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
5.  Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.** The committee that spent the most on Facebook advertising is Friends of Ashwani Jain.

```{r}
# copy data set
payee_media <- md_expenses |> 
# filter by expense category and name
  filter(
    expense_category == "Media"
  ) |> 
# get rid of NA in payee name
  drop_na(payee_name) |> 
# group by name
  group_by(payee_name) |> 
# arrange by name to check for NAs. hopefully i did it right
arrange(payee_name)

# write new csv
write_csv(payee_media, "data/media_expenses.csv")

# load new csv
payee_media_clean <- read_csv("data/media_expenses_clean.csv")

facebook_expenses <- payee_media_clean |> 
# filter by meta
  filter(payee_clean == "META PLATFORMS") |> 
# group by committee
  group_by(committee_name) |> 
# add up total amounts per committee
  summarize(
    total_amount = sum(amount)
  ) |> 
# arrange by top totals
arrange(desc(total_amount))

```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data%5Bdirection%5D=desc&sort_data%5Bmode%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** Most of the top spenders were committees supporting various Maryland gubernatorial candidates for the 2022 Democratic primary, such as Ashwani Jain (who spent the most), Peter Franchot and Jon Baron. Other candidates represented were for smaller or more local positions -- a notable one being Salimah Jasani, who was running for the Baltimore City Board of Ed. There were also two special interest PACs present -- CASA, which advocates for progressive policy on housing and immigration, and the Working Families Party, which also advocates for progressive policy, such as raising the minimum wage and environmental reform.

After reviewing Ashwani Jain's campaign ads on Facebook, I noticed that a majority of them were removed because they violated the platform's policy for political ads. Thus, I can't analyze their content or messaging. I also noticed that there was a huge discrepancy between the estimated audience size of these ads and their reported engagement. Although Jain's campaign aimed to reach upward to a million users, many of these ads were seen by a few thousand to tens of thousands of users. Jain's campaign also spent around \$400 per ad.

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.** Wes Moore and Aruna Miller, who won the Democratic primary, is missing from the Q1 list. Moore and Miller devoted most of their media expenses to SKDK, a political consulting firm that works with Democratic politicians. They also spent a lot of money on FogLamp, which is an internet system(?) and Apollo Artisty, a creative digital agency that works on branding, social engagement and content strategy.

```{r}
# check if moore or cox paid for facebook

search_fb_expenses <- payee_media_clean |> 
# filter by meta & committee name
  filter(
    payee_clean == "META PLATFORMS" &
    str_detect(committee_name, "Moore") |
    str_detect(committee_name, "Cox")
    ) |> 
# group by committee
  group_by(committee_name) |> 
# add up total amounts per committee
  summarize(
    total_amount = sum(amount)
  ) |> 
# arrange by top totals
arrange(desc(total_amount))

# assess moore media expenses
moore_expenses <- payee_media_clean |> 
# filter by moore committees
  filter(
    str_detect(committee_name, "Moore")
  ) |> 
# group by payee
  group_by(payee_clean) |> 
# sum up expenses by payee
  summarize(
    total_amount = sum(amount)
  ) |> 
arrange(desc(total_amount))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** Sorting by online advertising in a more general sense provided bigger names in Maryland politics, like Moore and Tom Perez. This shows that these politicians did use online advertising but not necessarily ads on Facebook. There are also more non-candidate committees listed, like the teachers/public educators' union, CASA, and the Maryland House Democratic Caucus.

The top five committees who spent the most on on Facebook ads from Q1 all appear to have spent all of their online advertising strategy on Facebook.

```{r}
# copy data set
online_ads_expenses <- payee_media_clean |> 
# filter by purpose
  filter(expense_purpose == "Online Advertising") |> 
# group by committee
  group_by(committee_name) |> 
# sum up expenses by committee
  summarize(
    total_amount = sum(amount)
  ) |> 
# show top totals
arrange(desc(total_amount))

# copy data set
online_ads_fb <- payee_media_clean |> 
# filter by online ad purpose and meta payee
  filter(
    expense_purpose == "Online Advertising" &
    payee_clean == "META PLATFORMS"
  ) |> 
# group by committee name
  group_by(committee_name) |> 
# sum up total amount
  summarize(
    total_amount = sum(amount)
  ) |> 
arrange(desc(total_amount))
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5** Tom Perez and Wes Moore, who were the most competitive gubernatorial candidates in the 2022 Democratic primary, did not rely on a media strategy centered on Facebook.

Most of the analysis in this lab focused on Facebook expenditures. I'd be interested in researching how consulting/PR companies (which Moore spent a lot of money on) utilize social media. I'd also like to compare how much was spent on other forms of advertising (print, broadcast, flyers, etc.) to online ads.

I think one story from this analysis can be looking into Facebook's evolving role in political campaigns from a local level. After looking up Ashwani Jain on Google I learned that he used to be an Obama administration official, and it seems like he took a lot of cues from this former president's campaign. I heard that the Obama campaign relied on social media (Facebook especially) to engage youth voters. From the numbers, it seems like this wasn't a successful strategy for Jain. Obama was in office over a decade ago now, and our relationship with and usage of Facebook has vastly changed since then.
